Compositional Quantitative Resource Analysis of C Programs

### 2014-06-25 paper/main.tex
trying to improve the introduction

Static analysis of quantitative properties of imperative programs is
an active research area and recent years have seen many innovations.
Notable tools that have been developed include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
%
While the these tools can derive impressive results for realistic
software, there still exist shortcomings that hamper the application
of the existing techniques in practice.
\begin{itemize}
\item Analysis tools are black boxes that either deliver a result or fail
  without enabling \emph{user interaction} to manually or
  semi-automatically derive bounds for challenging parts of the
  program.
\item The computed bounds are often \emph{non-compositional} local
  bounds (for a single (nested) loop) that are difficult to combine to
  global whole program bounds.
\item Existing techniques often rely on complex external tools such as
  abstract interpretation based invariant
  generation~\cite{GulwaniMC09} or translation of the program into a
  term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
  providing \emph{verifiable certificates} for the correctness of the
  derived bound.
\end{itemize}
%
While there has been much progress in static quantitative analysis,
Knuth correctly points out in a recent interview~\cite{KnuthInter}
that the state-of-the-art in formal quantitative methods still falls
short in comparison with semantic techniques.
%
\begin{quote}
  % Consider, for example, a computer program that sorts a list of
  % numbers into order. 
  $[\cdots]$ Thanks to the work of Floyd, Hoare, and others,
  we have formal definitions of semantics, and tools by which we can
  verify that sorting is indeed always achieved. My job is to go
  beyond correctness, to an analysis of such things as the program's
  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
  describes the program's performance, and all of my colleagues agree
  with me that the recurrence is "obviously" valid. Yet I have no
  formal tools by which I can prove that my recurrence is right. I
  don't really understand my reasoning processes at all!
  \vspace{-3ex}
  \begin{flushright}
    -- Donald E. Knuth, 2014
  \end{flushright}
\end{quote}
%
The goal of this work is to develop a resource analysis framework for
C programs that is based on a solid semantic foundation.  The choice
of C is primarily motivated by our ongoing work on the formally
verified hypervisor kernel CertiKOS~\cite{GuVFSC11}.  CertiKOS is
mainly developed in C and is supposed to provide verified guaranties
on timing and memory usage.  Moreover, C is also a natural choice
because it is still the most widely used language for system
development; in particular in safety-critical embedded and real-time
systems where resource bound analyses are often required by regulatory
authorities~\cite{May2013}.

In a recent work, we formally verified a bound on the stack usage of
CertiKOS

In this article, we present the first resource analysis framework for
C that makes it possible to combine automatically and interactively
derived resource bounds.  It works automatically, efficiently and
precisely for many examples, can be used interactively for manual
bound derivation if the automation fails, and produces verifiable
certificates for the bounds.  Our contributions are as follows.
%
\begin{enumerate}
\item We define a operational cost semantics for C programs that
  accounts a constant cost for each statement.  The cost is parametric
  in a cost metric and the semantics models cost of terminating and
  non-terminating executions.  A specific \emph{tick} function can be
  inserted in the source code to further customize the cost model.

\item We describe an automatic static analysis technique for deriving
  bounds for the cost that is defined by the operational semantics.
  It is based on amortized resource analysis for functional
  programs~\cite{Jost03,HoffmannAH12} and derives bounds that are
  non-negative linear combinations of sizes $|[x,y]|$ of intervals
  between program variables $x,y$.
\item We show with a publically available prototype implementation and
  an experimental evaluation that our novel amortized resource
  analysis works efficiently for realistic programs and that the
  derived constant factors in the bounds are close to or identical
  with the optimal ones.
\item We define a quantitative Hoare logic for interactively deriving
  cost bounds for C programs.  The logic is implemented and proved
  sound in the Coq Proof Assistant with respect to CompCert C
  light.  We show that our automatic analysis readily produces
  derivations in the quantitative Hoare logic.
% \item We prove the completeness of the program logic
%   with respect to the operational cost semantics.
\end{enumerate}
%
Our approach complements existing work since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  The quantitative Hoare logic and the automatic amortized
analysis enable semi-automatic interactive resource-bound development.
An automatically derived bound can be used in the quantitative logic
and a hand derived bound can be used by the automatic analysis.  Both
the automatic amortized analysis and the quantitative logic are
naturally compositional and describe the resource behavior of code
fragments without referring to the source code.  As a result, we
naturally derive global bounds that are functions of the input
parameters of the program.  To the best of our knowledge, our system
is the only one that can automatically handle resources like memory
that may become available during the execution.

>>>>

Static analysis of quantitative properties of imperative programs is
an active research area and recent years have seen many innovations.
Notable tools that have been developed include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
%
While the these tools can derive impressive results for realistic
software, there still exist shortcomings that hamper the application
of the existing techniques in practice.
\begin{itemize}
\item Analysis tools are black boxes that either deliver a result or fail
  without enabling \emph{user interaction} to manually or
  semi-automatically derive bounds for challenging parts of the
  program.
\item The computed bounds are often \emph{non-compositional} local
  bounds (for a single (nested) loop) that are difficult to combine to
  global whole program bounds.
\item Existing techniques often rely on complex external tools such as
  abstract interpretation based invariant
  generation~\cite{GulwaniMC09} or translation of the program into a
  term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
  providing \emph{verifiable certificates} for the correctness of the
  derived bound.
\end{itemize}
%
While there has been much progress in static quantitative analysis,
Knuth correctly points out in a recent interview~\cite{KnuthInter}
that the state-of-the-art in formal quantitative methods still falls
short in comparison with semantic techniques.
%
\begin{quote}
  % Consider, for example, a computer program that sorts a list of
  % numbers into order. 
  $[\cdots]$ Thanks to the work of Floyd, Hoare, and others,
  we have formal definitions of semantics, and tools by which we can
  verify that sorting is indeed always achieved. My job is to go
  beyond correctness, to an analysis of such things as the program's
  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
  describes the program's performance, and all of my colleagues agree
  with me that the recurrence is "obviously" valid. Yet I have no
  formal tools by which I can prove that my recurrence is right. I
  don't really understand my reasoning processes at all!
  \vspace{-3ex}
  \begin{flushright}
    -- Donald E. Knuth, 2014
  \end{flushright}
\end{quote}
%
In this work, we develop a resource analysis framework for C programs
that is based on a solid semantic foundation in which we start with a
formally defined cost model of C programs and derive resource bounds
together with easily verifiable certificates of their soundness with
respect to the cost model. The choice of C is primarily motivated by
our ongoing work on the formally verified hypervisor kernel
CertiKOS~\cite{GuVFSC11}.  CertiKOS is mainly developed in C and is
supposed to provide verified guaranties on timing and memory usage.
Moreover, C is also a natural choice because it is still the most
widely used language for system development; in particular in
safety-critical embedded and real-time systems where resource bound
analyses are often required by regulatory authorities~\cite{May2013}.

XXXXXXXXXX

Our starting point is a recent work~\cite{veristack14}, in which we
have formally verified bounds on the stack usage of C programs.  A key
part of the stack-bound verification is based on a quantitative Hoare
logic for deriving abstract stack bounds for CompCert Clight programs
that depend on the depth of (recursive) function calls.  For
non-recursive programs it is possible to compute derivations for such
stack bounds in the quantitative Hoare logic automatically.  By
compiling these bounds together with the C program, we have been able
to automatically derive stack bounds C programs without recursion such
as CertiKOS.

Our \emph{first contribution} is a generalization of the quantitative
Hoare logic for CompCert Clight that enables us to derive resource
bounds that are parametric in the resource of interest.  The
quantitative logic is implemented in the Coq Proof Assistant and
proved sound with respect to a small-step operational cost semantics
for Clight that counts a constant resource consumption for each
statement.  The resource consumption is defined by a cost metric and can be
negative to express the release of resources (e.g, as needed to reason
about memory usage).  The semantics models cost of terminating and
non-terminating executions and a specific \emph{tick} function can be
inserted in the source code to further customize the cost model.

- Problem: very tedious to reason about large programs
- automatic amortized analysis was inspiration for logic
- but hard to apply to C programs
- main problem: negative integers and precise bounds for \emph{for loops} like in python
  and control C-like control-flow with break and return

Our \emph{second contribution} is

In this article, we present the first resource analysis framework for
C that makes it possible to combine automatically and interactively
derived resource bounds.  It works automatically, efficiently and
precisely for many examples, can be used interactively for manual
bound derivation if the automation fails, and produces verifiable
certificates for the bounds.  Our contributions are as follows.
%
\begin{enumerate}
\item

\item We describe an automatic static analysis technique for deriving
  bounds for the cost that is defined by the operational semantics.
  It is based on amortized resource analysis for functional
  programs~\cite{Jost03,HoffmannAH12} and derives bounds that are
  non-negative linear combinations of sizes $|[x,y]|$ of intervals
  between program variables $x,y$.
\item We show with a publically available prototype implementation and
  an experimental evaluation that our novel amortized resource
  analysis works efficiently for realistic programs and that the
  derived constant factors in the bounds are close to or identical
  with the optimal ones.
\item We define a quantitative Hoare logic for interactively deriving
  cost bounds for C programs.  The logic is implemented and proved
  sound in the Coq Proof Assistant with respect to CompCert C
  light.  We show that our automatic analysis readily produces
  derivations in the quantitative Hoare logic.
% \item We prove the completeness of the program logic
%   with respect to the operational cost semantics.
\end{enumerate}

XXXXX

Our approach complements existing work since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  The quantitative Hoare logic and the automatic amortized
analysis enable semi-automatic interactive resource-bound development.
An automatically derived bound can be used in the quantitative logic
and a hand derived bound can be used by the automatic analysis.  Both
the automatic amortized analysis and the quantitative logic are
naturally compositional and describe the resource behavior of code
fragments without referring to the source code.  As a result, we
naturally derive global bounds that are functions of the input
parameters of the program.  To the best of our knowledge, our system
is the only one that can automatically handle resources like memory
that may become available during the execution.
